### 2017-05-07 2017mergingmerges/section/01-Introduction.tex
Fix sigma -> bs in advance rules
Cite sync kahn networks
edits and adjust page order

Suppose we have two input streams of numeric identifiers, and wish to perform some analysis on these identifiers. The identifiers from both streams are sorted, but may include duplicates. We wish to produce an output stream of unique identifiers from the first input stream, as well as produce the unique union of both streams. Here is how we might write the source code, where @S@ is for @S@-tream.
\begin{code}
  uniquesUnion : S Nat -> S Nat -> (S Nat, S Nat)
  uniquesUnion sIn1 sIn2
   = let  sUnique = group sIn1
          sMerged = merge sIn1 sIn2
          sUnion  = group sMerged
     in   (sUnique, sUnion)
\end{code}

@@

To implement this, we might write each operator as its own concurrent process, sending stream elements over intra-process channels. Developing this could be easy or hard, depending on the available language features for concurrency. However, the \emph{performance tuning} of such a system, such as whether we need back-pressure to prevent buffers from being overrun, or how to chunk stream data to reduce communication overhead, is invariably a headache.

% We might instead define a uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream. Each operator could be given this interface, so that values from each stream are computed on demand. This approach is commonly taken in database systems \cite{Graefe:Volcano}. However, this `pull' model does not support operators with multiple outputs, such as our example, at least not without unbounded buffering. Suppose a consumer pulls many elements from the @sUnique@ output stream. In order to perform the @group@ operation, the implementation needs to pull the corresponding source elements from the @sIn1@ input stream \emph{as well} as buffering an arbitrary number of them. It needs to buffer these elements because they are also needed to perform the @merge@ operation. When a consumer finally pulls from @sUnion@ we will be able to drain the buffer of @sIn1@ elements, but not before.

@@

% Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject operators with value dependent control flow such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.

% The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but is based around affine loops, which makes it difficult to support value-dependent operators  such as @group@ and @merge@.

@@

>>>>

Suppose we have two input streams of numeric identifiers, and wish to perform some analysis on them. The identifiers from both streams are sorted, but may include duplicates. We wish to produce an output stream of unique identifiers from the first input stream, as well as produce the unique union of both streams. Here is how we might write the source code, where @S@ is for @S@-tream.
\begin{code}
  uniquesUnion : S Nat -> S Nat -> (S Nat, S Nat)
  uniquesUnion sIn1 sIn2
   = let  sUnique = group sIn1
          sMerged = merge sIn1 sIn2
          sUnion  = group sMerged
     in   (sUnique, sUnion)
\end{code}

@@

To implement this program we might write each operator as its own concurrent process, sending stream elements over intra-process channels. Developing this could be easy or hard, depending on the available language features for concurrency. However, the \emph{performance tuning} of such a system, such as using back-pressure to prevent buffers from being overrun, or how to chunk stream data to amortize communication overhead, is invariably a headache.

@@

@@

% BL: This is now discussed in S2.5 Breaking It Down.
% We might instead define a uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream. Each operator could be given this interface, so that values from each stream are computed on demand. This approach is commonly taken in database systems \cite{Graefe:Volcano}. However, this `pull' model does not support operators with multiple outputs, such as our example, at least not without unbounded buffering. Suppose a consumer pulls many elements from the @sUnique@ output stream. In order to perform the @group@ operation, the implementation needs to pull the corresponding source elements from the @sIn1@ input stream \emph{as well} as buffering an arbitrary number of them. It needs to buffer these elements because they are also needed to perform the @merge@ operation. When a consumer finally pulls from @sUnion@ we will be able to drain the buffer of @sIn1@ elements, but not before.

% TODO: move to related work.
% The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but is based around affine loops, which makes it difficult to support value-dependent operators  such as @group@ and @merge@.

### 2017-05-07 2017mergingmerges/section/02-Processes.tex
edits and adjust page order

The definition of the @group@ combinator which detects groups of successive identical elements in the input stream is given in Figure~\ref{fig:Process:Group}. The process emits the first value pulled from the stream and every value that is different from the last one that was pulled. For example, when executed on the input stream @[1,2,2,3]@, the process will produce the output @[1,2,3]@. We include the concrete representation and a diagram of the process viewed as a state machine.

@@

The body of the combinator is a record that defines the process. The @ins@ field defines the set of input streams and the @outs@ field the set of output streams. The @heap@ field gives the initial values of each of the local variables. The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the initial instruction. In this form the output stream is defined via a parameter, rather than being the result of the combinator, as in the representation of @uniquesUnion@ from \S\ref{s:Introduction}.

@@

\subsection{Breaking It Down}
We started with a pure functional program as in \S\ref{s:Introduction}, reimagined it as a dataflow graph, then interleaved imperative code that implemented two of the operators in that dataflow graph. Importantly, we needed to \emph{break down} the definition of each of the operators into imperative statements so that we could interleave their execution appropriately. We do this because the standard, single-threaded evaluation semantics of functional programs does not allow us to evaluate stream programs that contain both splits and joins in a space efficient way. Returning to the definition of @uniquesUnion@ from \S\ref{s:Introduction}, we cannot simply execute the @group@ operator on its entire input @sIn1@ before executing the @merge@ operator, as that would require us to buffer all data read from @sIn1@. Instead, in the fused result program we have performed the job of a concurrent scheduler at compile time. The flow of control alternates between the implementation of each of the operators, but as the instructions are interleaved directly there is no overhead due to context switching.

>>>>

The definition of the @group@ combinator which detects groups of successive identical elements in the input stream is given in Figure~\ref{fig:Process:Group}. The process emits the first value pulled from the stream and every value that is different from the last one that was pulled. For example, when executed on the input stream @[1,2,2,3]@, the process will produce the output @[1,2,3]@. We include the concrete representation and a diagram of the process when viewed as a state machine.

@@

The body of the combinator is a record that defines the process. The @ins@ field defines the set of input streams and the @outs@ field the set of output streams. The @heap@ field gives the initial values of each of the local variables. The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the initial instruction. In this form, the output stream is defined via a parameter, rather than being the result of the combinator, as in the representation of @uniquesUnion@ from \S\ref{s:Introduction}.

@@

% -----------------------------------------------------------------------------
\subsection{Breaking It Down}
We started with a pure functional program as in \S\ref{s:Introduction}, reimagined it as a dataflow graph, then interleaved imperative code that implemented two of the operators in that dataflow graph. Importantly, we needed to \emph{break down} the definition of each of the operators into imperative statements so that we could interleave their execution appropriately. We do this because the standard, single-threaded evaluation semantics of functional programs does not allow us to evaluate stream programs that contain both splits and joins in a space efficient way. Returning to the definition of @uniquesUnion@ from \S\ref{s:Introduction}, we cannot simply execute the @group@ operator on its entire input @sIn1@ before executing the @merge@ operator, as that would require us to buffer all data read from @sIn1@. Instead, in the fused result program we have performed the job of a concurrent scheduler at compile time. The flow of control alternates between the implementation of each of the operators, but as the instructions are interleaved directly there is no overhead due to context switching.
